# In Figure 1, we show a PCA plot of Principal Components 1 and 2 (PC1,PC2) from the complete patient gene expression dataset. This dataset consists of gene expression profiles from 87 patients having either a good or a poor prognosis. We note that there is closer or more discrete clustering among those patients with a poor prognosis than than those with a good prognosis, with some overlap. Figure 2 shows the fraction each principal component contributes to the variance explained by the dataset. PC1 and PC2 contribute to a little over 30% of the variation. 

#In order to produce a Hopfield network for each gene of the 30 genes assessed in the entire dataset, we calculate the correlation values between a gene and it's correlation with all of the other 29 genes, one pair at a time. Figure 3 shows the distribution of the standardized (to a standard normal distribution) correlation values for each pair derived from a 30 x 30 matrix. A minor peak above 3 is seen, indicating the correlation of a gene with itself. Figures 4 and 5 show the distribution of the standardized correlation values for each gene pair in the Poor and Good prognosis dataset, respectively. Again, we note a minor peak of 1 on the x-axis,representing the correlation values between a gene and itself.

```{r Load the data and parse appropriately, echo=FALSE, message=FALSE, results='hide'}
library(tidyverse)
library(plotrix)
dat = read_csv("HW2_data.csv")

dat[dat$`prognosis` == 'GOOD', ][, "prognosis"] <- 2
dat[dat$`prognosis` == 'POOR', ][, "prognosis"] <- 1

# Calculate Z-scores for each probe / patient


datP = dat %>% filter(prognosis==1) %>% drop_na()
datG = dat %>% filter(prognosis==2) %>% drop_na()
```

```{r Run a PCA on the dataset, echo=FALSE,message=FALSE,warning=FALSE}
group <- as.numeric(dat$prognosis)
x <- dat
pc <- prcomp(data.matrix(x))
# ?prcomp
#names(pc)
plot(pc$x[, 1], pc$x[, 2], col = group, xlab = "PC1", ylab = "PC2",sub="Figure 1. PCA of Expression Dataset in 87 Patients with Either a Good or Poor Prognosis")
legend("topright", legend=c("GOOD", "POOR"),
       col=c("red", "black"), lty=1,box.lty=0)

# Plot the variance explained by each principal component
x=data.matrix(x)
cx <- sweep(x, 2, colMeans(x), "-")
sv <- svd(cx)
plot(sv$d^2/sum(sv$d^2), xlim = c(0, 15), type = "b", pch = 16, xlab = "principal components", 
    ylab = "variance explained",sub="Figure 2. Plot of Fractions Each Principal Component Contributes to the Variance of the Dataset")

```


```{r Perform the correlation assessment on the whole dataset, echo=FALSE,message=FALSE}
core=cor(data.matrix(dat[,1:30])) # correlations

#scale(core,center=TRUE,scale=TRUE) # (z-score calculation-ish?) subtract the mean of (each probe's correlation to every other probe) and divide by the standard deviation # row wise caluclation 
sdn=(core-mean(core))/sd(core) # (z-score calculation-ish?) subtract the mean of (each probe's correlation to every other probe) and divide by the standard deviation #Column wise calculation
plot(density(data.matrix(sdn)),main="",sub="Figure 3. Distribution of Gene-to-Gene Correlation Values Across All 87 Patients") # distribution

write.table(sdn, file = "zscore.txt", sep = "\t",row.names = FALSE, col.names=FALSE) # remove row names
```


```{r Perform the correlation assessment on the POOR dataset, echo=FALSE,message=FALSE,warning=FALSE}
drop=c("X37439_at","prognosis","X31989_s_at")
core=cor(data.matrix(datP)) %>% data.frame() %>% slice( 1:(n()-2)) %>% select(-drop) %>% na.omit()
sdn=(data.matrix(core)-mean(data.matrix(core)))/sd(data.matrix(core))
plot(density(data.matrix(core)),main="",sub="Figure 4. Distribution of Gene-to-Gene Correlation Values Across Patients with a Poor Prognosis") # distribution)

```

```{r Perform the correlation assessment on the GOOD dataset, echo=FALSE,message=FALSE,warning=FALSE}
drop="prognosis"
core=cor(data.matrix(datG)) %>% data.frame() %>% slice( 1:(n()-2)) %>% select(-drop) %>% na.omit()
sdn=(data.matrix(core)-mean(data.matrix(core)))/sd(data.matrix(core))
plot(density(data.matrix(core)),main="",sub="Figure 5. Distribution of Gene-to-Gene Correlation Values Across Patients with a Good Prognosis") # distribution

```

